Reduced Nucleotide Diversity on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-10-27 manuscript.tex
edits

\begin{abstract}
X and Y chromosomes differ in effective population size ($N_{e}$), rates of recombination, and exposure to natural selection, all of which can significantly affect levels of genetic variability within the genome. On Y chromosomes with suppressed recombination, selection can remove genetically-linked neutral variation, resulting in reduced $N_{e}$ and the level of polymorphism on Y compared to X chromosomes or autosomes. However, non-selective factors including female biased sex ratios and high variance in male fitness can also reduce Y-linked $N_{e}$, making it difficult to infer the causes of low Y-chromosome diversity. Here, we investigate the factors affecting X- and Y-linked polymorphism during the early stages of plant sex chromosome evolution in \textit{Rumex hastatulus} (Polygonaceae). Strikingly, we find that nucleotide diversity on the Y is more than 40-fold lower than on the X, suggesting an extensive reduction in diversity. We demonstrate that the magnitude of diversity loss is inconsistent with a purely neutral model that accounts for the occurrence of female-biased sex ratios in this species, or by reduced male $N_{e}$ caused by high variance in male fitness. Rather, using forward simulations and population genetic estimates of the distribution of fitness effects of deleterious mutations, we show that Y diversity levels can be explained by the action of linked selection against deleterious mutations, although selective sweeps may also be playing a role. Given the recent origin of \textit{R. hastatulus} sex chromosomes, our study is consistent with the prediction that the early stages of Y chromosome degeneration are characterized by strong effects of selective interference experienced by a large number of selected sites. These results suggest that selective interference is playing a major role in the ongoing degeneration of the Y chromosome, rather than simply neutral drift following gene silencing. 
\end{abstract}

@@

Several demographic factors are expected to modulate or accentuate differences in neutral polymorphism maintained on sex chromosomes \citep{ellegren2009}. These include population subdivision and sex-biased dispersal, deviations from a 1:1 breeding sex ratio, and high sex-specific variance in reproductive success \citep{caballero1995,charlesworth2001,laporte2002,pool2007}. In theory, these processes all result in asymmetries in female vs male $N_{e}$, that could lead to different levels of neutral polymorphism on the sex chromosomes than are expected based on the autosomes (\citep{pool2007}). As evidence for this, coalescent simulations have shown that models of male-biased migration during human out-of-africa dispersal result in levels of X/A neutral diversity similar to what is observed \citep{keinan2009} (though it is worth noting that there is considerable variation in current estimates of X/A in humans, and other demographic scenarios have been suggested \citep{hammer2010,bustamante2009}). In addition, disproportionate losses of sex-linked compared to autosomal variation arising from population bottleneck events have been inferred in chimpanzees and orangutans \citep{kaessmann2001,fischer2006}, the house mouse \textit{Mus musculus} \citep{baines2007}, and Drosophila \citep{andolfatto2001}, highlighting the importance of demography in affecting patterns of sex-linked and autosomal neutral variation.

In addition to demographic factors, evolutionary theory predicts that neutral variability on Y chromosomes with suppressed recombination can be significantly reduced by Hill-Robertson Interference (HRI), which causes a reduction in $N_{e}$ of neutral sites that are genetically linked to sites experiencing selection. The major types of HRI include (i) selection on strongly beneficial mutations ("selective sweeps") \citep{smith1974hitch,aquadro1994} (ii) selection against strongly deleterious mutations
) \citep{charlesworth1996background,charlesworth1994effect} \citep{charlesworth1996CB,charlesworth2000degeneration}, and (iii) intereference between weakly selected sites (weak selection Hill-Robertson interference). In all of these processes, linkage  disequilibrium between selected and neutral sites results in a level of neutral variation below the level predicted in the absence of selection. In a large nonrecombining region experiencing sufficiently high deleterious mutation rates, the distinction between the effects of strongly- and weakly- selected mutations begins to break down, and have been explored primarily using forward computer simulations (refs).

Consistent with an important role for Hill-Robertson interference in driving down diversity on the Y chromosome, studies examining Y chromosome variability in mammals \citep{hellborg2004,bachtrog2013NRG,Wilsonsayres2014} and Drosophila \citep{mcallister1999,bachtrog2000} have revealed that Y-linked diversity is considerably lower than predicted under models of neutral evolution, with purifying selection playing a key role.

@@

\subsection*{Estimating nucleotide diversity on sex chromosomes and autosomes}
For each locus in our analysis, we calculated Watterson’s (1975) estimator of the population parameter $\theta=4N_{e}\mu$, where $N_{e}$ is the effective population size, and $\mu$ is the mutation rate \citep{watterson1975}, using a modified version of the Perl program Polymorphurama \citep{bachtrog2006}. To compare sex-linked and autosomal loci, we calculated the average value of $\theta$, weighted by the number of synonymous sites in each gene (\X Figure 2; Table 1). We obtained 95 percent confidence intervals for our estimates of the X/A and Y/A diversity ratios by bootstrapping by gene using the BCa method \citep{efron1994} implemented in the Boot package in R \citep{canty2012boot}, and calculating X/A and Y/A diversity on each iteration for 20000 replicates each. Bootstrapping was conducted on the final filtered set of 173 sex-linked, and 12355 autosomal genes from the Texas race, and separately for the 176 sex-linked and 11349 autosomal genes from the North Carolina race. Note that the lack of recombination on the Y chromosome implies that assumptions about independence across loci are likely violated, suggesting that the true uncertainty in our Y/A diversity estimate is likely an underestimate. To address this issue, we used a maximum likelihood approach, implemented in a modified version of the MLHKA software \citep{wright2004hka} to independently estimate a credibility interval for the Y/A ratio (Figure S1). Because of the thousands of genes involved, a likelihood method incorporating divergence to control for heterogeneity in mutation rate is not feasible, since this would involve maximizing the likelihood estimate of the mutation rate for each locus independently.  Therefore, we applied a method that assumed no heterogeneity in mutation rate, no recombination between Y-linked genes, and free recombination between autosomal loci. Our model therefore had two parameters; theta for autosomal genes and the Y/A ratio. We varied both parameters across a grid of points and evaluated the likelihood, where the Y/A ratio varied from 0.001 to 1, and Watterson's theta on the autosomes varied from 0.001 to 0.01. We also tested whether estimates of diversity for the X chromosome calculated from phased sequences from females were consistent with estimates from phased sequences from males. As no significant difference was observed (\X Figure? Table?), we report only results from females.

Finally, we tested for population substructure within and between the two sex chromosome races to control for the possibility that hidden substructure in our sampled populations could affect estimates of diversity. To do this, we constructed neighbor-joining trees of sex-linked sequences from all populations in our study. The analysis was conducted on an alignment of the X- and Y-linked genes from \textit{R. hastatulus}, with orthologous autosomal sequences from the non-dioecious but closely related outgroup species \textit{R. bucephalophorus} used to root the tree (Figure S2). We used the Neighbor-Joining method \citep{saitou1987neighbor}, with evolutionary distances computed using Maximum Composite Likelihood \citep{tamura2011mega5}. The inferred trees revealed strong support for Y-linked genes from the $XY_{1}XY_{2}$ ("North Carolina") race being paraphyletic (98\% bootstrap support), with samples from two populations (hereafter the "SC sub-clade") forming a monophyletic group, and samples from Florida and Georgia (hereafter the "FL sub-clade") also forming a monophyletic group, that was more closely related to the XY ("Texas") race (Figure S2). In light of this evidence for significant population substructure, and to consider between population differences, we estimated X and Y diversity in each of the three sub-clades (SC, FL, TX) separately.

>>>>

\begin{abstract}
X and Y chromosomes differ in effective population size ($N_{e}$), rates of recombination, and exposure to natural selection, all of which can significantly affect levels of genetic variability within the genome. On Y chromosomes with suppressed recombination, selection can remove genetically-linked neutral variation, resulting in reduced $N_{e}$ and the level of polymorphism on Y compared to X chromosomes or autosomes. However, non-selective factors including female biased sex ratios and high variance in male reproductive success can also reduce Y-linked $N_{e}$, making it difficult to infer the causes of low Y-chromosome diversity. Here, we investigate the factors affecting X- and Y-linked polymorphism during the early stages of plant sex chromosome evolution in \textit{Rumex hastatulus} (Polygonaceae). Strikingly, we find that nucleotide diversity on the Y is more than 40-fold lower than on the X, suggesting an extensive reduction in diversity. We demonstrate that the magnitude of diversity loss is inconsistent with a purely neutral model that accounts for the occurrence of female-biased sex ratios in this species, or by reduced male $N_{e}$ caused by high variance in male fitness. Rather, using forward simulations and population genetic estimates of the distribution of fitness effects of deleterious mutations, we show that Y diversity levels can be explained by the action of linked selection against deleterious mutations, although selective sweeps may also be playing a role. Given the recent origin of \textit{R. hastatulus} sex chromosomes, our study is consistent with the prediction that the early stages of Y chromosome degeneration are characterized by strong effects of selective interference experienced by a large number of selected sites, rather than simply neutral drift following gene silencing. 
\end{abstract}

@@

Several demographic factors are expected to modulate or accentuate differences in neutral polymorphism maintained on sex chromosomes \citep{ellegren2009}. These include population subdivision and sex-biased dispersal, deviations from a 1:1 breeding sex ratio, and high sex-specific variance in reproductive success \citep{caballero1995,charlesworth2001,laporte2002,pool2007}. In theory, these processes all result in asymmetries in female vs male $N_{e}$, that could lead to different levels of neutral polymorphism on the sex chromosomes than are expected based on the autosomes (\citep{pool2007}). As evidence for this, coalescent simulations have shown that models of male-biased migration during human out-of-Africa dispersal result in levels of X/A neutral diversity similar to what is observed \citep{keinan2009} (though it is worth noting that there is considerable variation in current estimates of X/A in humans, and other demographic scenarios have been suggested \citep{hammer2010,bustamante2009}). In addition, disproportionate losses of sex-linked compared to autosomal variation arising from population bottleneck events have been inferred in chimpanzees and orangutans \citep{kaessmann2001,fischer2006}, the house mouse \textit{Mus musculus} \citep{baines2007}, and Drosophila \citep{andolfatto2001}, highlighting the importance of demography in affecting patterns of sex-linked and autosomal neutral variation.

In addition to demographic factors, evolutionary theory predicts that neutral variability on Y chromosomes with suppressed recombination can be significantly reduced by Hill-Robertson Interference (HRI), which causes a reduction in $N_{e}$ of neutral sites that are genetically linked to sites experiencing selection. The major types of HRI include (i) selection on strongly beneficial mutations ("selective sweeps") \citep{smith1974hitch,aquadro1994} (ii) selection against strongly deleterious mutations
) \citep{charlesworth1996background,charlesworth1994effect} \citep{charlesworth1996CB,charlesworth2000degeneration}, and (iii) intereference between weakly selected sites (weak selection Hill-Robertson interference). In all of these processes, linkage  disequilibrium between selected and neutral sites results in a level of neutral variation below the level predicted in the absence of selection.

Consistent with an important role for Hill-Robertson interference in driving down diversity on the Y chromosome, studies examining Y chromosome variability in mammals \citep{hellborg2004,bachtrog2013NRG,Wilsonsayres2014} and Drosophila \citep{mcallister1999,bachtrog2000} have revealed that Y-linked diversity is considerably lower than predicted under models of neutral evolution, with linked selection playing a key role.

@@

\subsection*{Estimating nucleotide diversity on sex chromosomes and autosomes}
For each locus in our analysis, we calculated Watterson’s (1975) estimator of the population parameter $\theta=4N_{e}\mu$, where $N_{e}$ is the effective population size, and $\mu$ is the mutation rate \citep{watterson1975}, using a modified version of the Perl program Polymorphurama \citep{bachtrog2006}. To compare sex-linked and autosomal loci, we calculated the average value of $\theta$, weighted by the number of synonymous sites in each gene (\X Figure 2; Table 1). We obtained 95 percent confidence intervals for our estimates of the X/A and Y/A diversity ratios by bootstrapping by gene using the BCa method \citep{efron1994} implemented in the Boot package in R \citep{canty2012boot}, and calculating X/A and Y/A diversity on each iteration for 20000 replicates each. Bootstrapping was conducted on the final filtered set of 173 sex-linked, and 12355 autosomal genes from the Texas race, and separately for the 176 sex-linked and 11349 autosomal genes from the North Carolina race. Note that the lack of recombination on the Y chromosome implies that assumptions about independence across loci are likely violated, suggesting that the true uncertainty in our Y/A diversity estimate may be wider than implied by bootstrapping across genes. To address this issue, we also used a maximum likelihood approach, implemented in a modified version of the MLHKA software \citep{wright2004hka} to independently estimate a credibility interval for the Y/A ratio (Figure S1). Because of the thousands of genes involved, a likelihood method incorporating divergence to control for heterogeneity in mutation rate is not feasible, since this would involve maximizing the likelihood estimate of the mutation rate for each locus independently.  Therefore, we applied a method that assumed no heterogeneity in mutation rate, no recombination between Y-linked genes, and free recombination between autosomal loci. Our model therefore had two parameters; theta for autosomal genes and the Y/A ratio. We varied both parameters across a grid of points and evaluated the likelihood, where the Y/A ratio varied from 0.001 to 1, and Watterson's theta on the autosomes varied from 0.001 to 0.01. We also tested whether estimates of diversity for the X chromosome calculated from phased sequences from females were consistent with estimates from phased sequences from males. As no significant difference was observed (\X Figure? Table?), we report only results from females.

Finally, we tested for population substructure within and between the two sex chromosome races to control for the possibility that hidden substructure in our sampled populations could affect estimates of diversity. To do this, we constructed neighbor-joining trees of sex-linked sequences from all populations in our study. The analysis was conducted on an alignment of the X- and Y-linked genes from \textit{R. hastatulus}, with orthologous autosomal sequences from the non-dioecious but closely related outgroup species \textit{R. bucephalophorus} used to root the tree (Figure S2). We used the Neighbor-Joining method \citep{saitou1987neighbor}, with evolutionary distances computed using Maximum Composite Likelihood \citep{tamura2011mega5}. The inferred trees revealed strong support for Y-linked genes from the $XY_{1}XY_{2}$ ("North Carolina") race being paraphyletic (98\% bootstrap support), with samples from two populations (hereafter the "SC sub-clade") forming a monophyletic group, and samples from Florida and Georgia (hereafter the "FL sub-clade") also forming a monophyletic group, that was more closely related to the XY ("Texas") race (Figure S2). In light of this evidence for significant population substructure, and to consider between-population differences, we estimated X and Y diversity in each of the three sub-clades (SC, FL, TX) separately.
